---
title: 'EDS 241: Assignment 3'
author: "Paloma Cartwright"
date: "`r Sys.Date()`"
output: 
    pdf_document:
    toc: false
    number_sections: yes
header-includes:
  - \setlength{\parindent}{1em}
  - \usepackage{float}
--- 
  
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, 
                      echo = TRUE, message = FALSE, warning = FALSE)

# load packages
librarian::shelf("stargazer", "here", "tidyr", "tidyverse", "dplyr","stringr", "janitor", "huxtable",
           "cowplot", "ggplot2", "tinytex", "datasets", "tibble", "readxl", "estimatr", "patchwork",
           "car")


```

\noindent The goal is to estimate the causal effect of maternal smoking during pregnancy on infant birth weight using the treatment ignorability assumptions. The data are taken from the National Natality Detail Files, and the extract “SMOKING_EDS241.csv”' is a random sample of all births in Pennsylvania during 1989-1991. Each observation is a mother-infant pair. The key variables are:

**The outcome and treatment variables:**
`birthwgt` = birth weight of infant in grams
`tobacco` = indicator for maternal smoking

**The control variables:**
`mage` (mother's age) 
`meduc` (mother's education)
`mblack` (=1 if mother black)
`alcohol` (=1 if consumed alcohol during pregnancy)
`first` (=1 if first child), `diabete` (=1 if mother diabetic)
`anemia` (=1 if mother anemic)

# Clean and plot data

\noindent The following code loads the data.

```{r, include=TRUE}
# Load data
data <- read_csv(here("SMOKING_EDS241.csv"))

```

# Question 1

(a) What is the unadjusted mean difference in birth weight of infants with smoking and nonsmoking mothers? Under what assumption does this correspond to the average treatment effect of maternal smoking during pregnancy on infant birth weight? Provide some simple empirical evidence for or against this assumption.

```{r, include=TRUE}
smoker <- data %>% filter(tobacco == 1)
nonsmoker <- data %>% filter(tobacco == 0)

mean_smoker <- round(mean(smoker$birthwgt), 3)
mean_nonsmoker <- round(mean(nonsmoker$birthwgt), 3)

diff <- mean_nonsmoker - mean_smoker

```

The unadjusted mean difference in birth weight in grams for children whose mothers did not smoke versus those that did is `r diff` grams.

The assumption is that smoking status is independent of all other variables implying unconditional treatment ignorability. This would mean that smoking mothers and non-smoking mothers are completely randomly assigned and no other variables have an impact on whether mother's smoke or not. 


```{r, warning=F}
model1 <- lm(mblack ~ tobacco, data = data)

se_model = starprep(model1, stat = c("std.error"), se_type = "HC2", alpha = 0.05)
stargazer(model1, se = se_model, type="text")

```

The p-value for the impact of whether the mother is black on smoking during pregnancy is `r model1$p.value[2]`. This is less than 0.5 meaning that it is statistically significant and the race of the mother is correlated with whether they used tobacco during pregnancy. This would contradict the assumption that smoking status is independent and that there is unconditional treatment ignorability. 

# Question 2 

(b) Assume that maternal smoking is randomly assigned conditional on the observable covariates listed above. Estimate the effect of maternal smoking on birth weight using a linear regression. Report the estimated coefficient on tobacco and its standard error.

```{r, warning=FALSE} 
model2 <- lm_robust(birthwgt ~ tobacco + mage + meduc + mblack + alcohol + first + diabete + anemia,
                    data = data)

summary(model2)

```
The coefficient for the average impact of tobacco on birth weight in grams is `r model2$coefficients[2]` and the standard error is `r model2$std.error[2]`.

(c) Use the exact matching estimator to estimate the effect of maternal smoking on birth weight. 
For simplicity, consider the following covariates in your matching estimator: create a 0-1 indicator for mother's age (=1 if mage>=34), and a 0-1 indicator for mother's education (1 if meduc>=16), mother's race (mblack), and alcohol consumption indicator (alcohol). 
These 4 covariates will create 2*2*2*2 = 16 cells. 
Report the estimated average treatment effect of smoking on birthweight using the exact matching estimator and its linear regression analogue (Lecture 6, slides 12-14).

```{r}
data_matching <- data %>% 
  mutate(
    mage_sq = (mage*mage),
    mage = case_when(
      mage >= 34 ~ 1,
      mage <34 ~ 0),
    meduc = case_when(
      meduc >= 16 ~ 1, 
      meduc < 16 ~ 0
    ), 
    mblack = as.factor(mblack), 
    alcohol = as.factor(alcohol),
    g = paste0(mage, meduc, mblack, alcohol)
    )
```

```{r}
TIA_table <- data_matching %>%
  group_by(g, tobacco)%>% 
  summarise(n_obs = n(),
            bwgt_mean= mean(birthwgt, na.rm = T)) %>% #Calculate number of observations and Y mean by X by treatment cells
  gather(variables, values, n_obs:bwgt_mean) %>% #Reshape data
  mutate(variables = paste0(variables, "_", tobacco, sep=""))%>% #Combine the treatment and variables for reshaping
  pivot_wider(id_cols = g, names_from = variables, values_from = values) %>% #Reshape data by treatment and X cell
  ungroup() %>%  #Ungroup from X values
  mutate(bwgt_diff = bwgt_mean_1 - bwgt_mean_0, #calculate Y_diff
         w_ATE = (n_obs_0 + n_obs_1) / (sum(n_obs_0) + sum(n_obs_1)),
         w_ATT = n_obs_1 / sum(n_obs_1)) %>% #calculate weights
  mutate_if(is.numeric, round, 2) #Round data

stargazer(TIA_table, type= "text", summary = FALSE, digits = 2)

# MULTIVARIATE MATCHING ESTIMATE OF ATE
ATE=sum((TIA_table$w_ATE)*(TIA_table$bwgt_diff))
ATE

```

The exact matching estimator has an average treatment effect of `r ATE`.

```{r}

model3 <- lm(birthwgt ~ tobacco + 
                      mage + meduc + mblack + alcohol + 
                      mage:meduc + mage:mblack + mage:alcohol +
                      meduc:mblack + meduc:alcohol + mblack:alcohol +
                      mage:meduc:mblack + mage:meduc:alcohol + meduc:mblack:alcohol +
                      mage:meduc:mblack:alcohol, data = data_matching)

se_model = starprep(model3, stat = c("std.error"), se_type = "HC2", alpha = 0.05)
stargazer(model3, se = se_model, type="text")

```

(d) Estimate the propensity score for maternal smoking using a logit estimator and based on the following specification: mother’s age, mother’s age squared, mother’s education, and indicators for mother’s race, and alcohol consumption. 

```{r}
# this is the model of the propensity score 
ps_model <- glm(tobacco ~ mage + mage_sq + meduc + mblack + alcohol, family = binomial(), data = data_matching)
summary(ps_model)

EPS <- predict(ps_model, type = "response")
PS_WGT <- (data_matching$tobacco/EPS) + ((1-data_matching$tobacco)/(1-EPS))

head(PS_WGT)

```

(e) Use the propensity score weighted regression (WLS) to estimate the effect of maternal smoking on birth weight.

```{r}

# WLS USING EPS WEIGHTS
wls1 <- lm_robust(birthwgt ~ tobacco, data_matching, weights = PS_WGT)
summary(wls1)

wls2 <- lm_robust(birthwgt ~ tobacco + mage + mage_sq + meduc + mblack + alcohol, data_matching, weights = PS_WGT)
summary(wls2)
```

