---
title: 'EDS 241: Assignment 3'
author: "Paloma Cartwright"
date: "`r Sys.Date()`"
output: 
    pdf_document:
    toc: false
    number_sections: yes
header-includes:
  - \setlength{\parindent}{1em}
  - \usepackage{float}
--- 
  
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, 
                      echo = TRUE, message = FALSE, warning = FALSE)

# load packages
librarian::shelf("stargazer", "here", "tidyr", "tidyverse", "dplyr","stringr", "janitor", "huxtable",
           "cowplot", "ggplot2", "tinytex", "datasets", "tibble", "readxl", "estimatr", "patchwork",
           "car")


```

\noindent The goal is to estimate the causal effect of maternal smoking during pregnancy on infant birth weight using the treatment ignorability assumptions. The data are taken from the National Natality Detail Files, and the extract “SMOKING_EDS241.csv”' is a random sample of all births in Pennsylvania during 1989-1991. Each observation is a mother-infant pair. The key variables are:

**The outcome and treatment variables:**
`birthwgt` = birth weight of infant in grams
`tobacco` = indicator for maternal smoking

**The control variables:**
`mage` (mother's age) 
`meduc` (mother's education)
`mblack` (=1 if mother black)
`alcohol` (=1 if consumed alcohol during pregnancy)
`first` (=1 if first child), `diabete` (=1 if mother diabetic)
`anemia` (=1 if mother anemic)

# Clean and plot data

\noindent The following code loads the data.

```{r , include=TRUE}
# Load data
data <- read_csv(here("SMOKING_EDS241.csv"))

```
# Question 1

(a) What is the unadjusted mean difference in birth weight of infants with smoking and nonsmoking mothers? Under what assumption does this correspond to the average treatment effect of maternal smoking during pregnancy on infant birth weight? Provide some simple empirical evidence for or against this assumption.

```{r , include=TRUE}
smoker <- data %>% filter(tobacco == 1)
nonsmoker <- data %>% filter(tobacco == 0)

mean_smoker <- round(mean(smoker$birthwgt), 3)
mean_nonsmoker <- round(mean(nonsmoker$birthwgt), 3)

diff <- mean_nonsmoker - mean_smoker

```

The unadjusted mean difference in birth weight in grams for children whose mothers did not smoke versus those that did is `r diff` grams.

The assumption is that smoking status is independent of y(0) and y(1) implying unconditional treatment ignorability. 

**OH NOTES**
if you have income on lhs, whether you are a smoker or not will not be statistically different. Maybe the hypothesis is that you're testing whether the treatment status is independent of y1 y0
What would need to be true for smoker vs nonsmoker to be the treatment effect? We have treatment ignorability. We should be working toward conditional treatment ignorability. 

the hypothesis/assumpiton smoking status is independent of y(1) y(0) - unconditional treatment ignorability. We would need smoking status to be randomly assigned to mothers unconditionally. 

Run a model to provide evidence 
run income on smoking status, education on smoking status, age on smoking status - provide someeee evidence that smoking status is correlated with these variables. 
You can do this with the mean differences between them but you want to know if they are statistically different. running the regression gives you a test statistic already. if it were true that smoking was randomly assigned then the regressions of other vars on smoking would not be statistically significant. 
```{r, warning=F}
model1 <- lm_robust(mblack ~ tobacco, data = data)
summary(model1) 

```

The p-value for the impact of whether the mother is black on smoking during pregnancy is `r model1$p.value[2]`. This is less than 0.5 meaning that it is statistically significant and the race of the mother is correlated with whether they used tobacco during pregnancy. This would contradict the assumption that smoking status is independent and that there is unconditional treatment ignorability. 

# Question 2 

(b) Assume that maternal smoking is randomly assigned conditional on the observable covariates listed above. Estimate the effect of maternal smoking on birth weight using a linear regression. Report the estimated coefficient on tobacco and its standard error.

```{r, warning=FALSE} 
model2 <- lm_robust(birthwgt ~ tobacco + mage + meduc + mblack + alcohol + first + diabete + anemia,
                    data = data)
summary(model2)
```
The coefficient for the average impact of tobacco on birth weight in grams is `r model2$coefficients[2]` and the standard error is `r model2$std.error[2]`.

(c) Use the exact matching estimator to estimate the effect of maternal smoking on birth weight. 
For simplicity, consider the following covariates in your matching estimator: create a 0-1 indicator for mother's age (=1 if mage>=34), and a 0-1 indicator for mother's education (1 if meduc>=16), mother's race (mblack), and alcohol consumption indicator (alcohol). 
These 4 covariates will create 2*2*2*2 = 16 cells. 
Report the estimated average treatment effect of smoking on birthweight using the exact matching estimator and its linear regression analogue (Lecture 6, slides 12-14).

```{r}
data_matching <- data %>% 
  mutate(
    mage_sq = (mage*mage),
    mage = case_when(
      mage >= 34 ~ 1,
      mage <34 ~ 0),
    meduc = case_when(
      meduc >= 16 ~ 1, 
      meduc < 16 ~ 0
    ), 
    mblack = as.factor(mblack), 
    alcohol = as.factor(alcohol),
    g = paste0(mage, meduc, mblack, alcohol)
    )
```

```{r}
TIA_table <- data_matching %>%
  group_by(g, tobacco)%>% 
  summarise(n_obs = n(),
            bwgt_mean= mean(birthwgt, na.rm = T)) %>% #Calculate number of observations and Y mean by X by treatment cells
  gather(variables, values, n_obs:bwgt_mean) %>% #Reshape data
  mutate(variables = paste0(variables, "_", tobacco, sep=""))%>% #Combine the treatment and variables for reshaping
  pivot_wider(id_cols = g, names_from = variables, values_from = values) %>% #Reshape data by treatment and X cell
  ungroup() %>%  #Ungroup from X values
  mutate(bwgt_diff = bwgt_mean_1 - bwgt_mean_0, #calculate Y_diff
         w_ATE = (n_obs_0 + n_obs_1) / (sum(n_obs_0) + sum(n_obs_1)),
         w_ATT = n_obs_1 / sum(n_obs_1)) %>% #calculate weights
  mutate_if(is.numeric, round, 2) #Round data


stargazer(TIA_table, type= "text", summary = FALSE, digits = 2)

# MULTIVARIATE MATCHING ESTIMATES OF ATE AND ATT
ATE=sum((TIA_table$w_ATE)*(TIA_table$bwgt_diff))
ATE
ATT=sum((TIA_table$w_ATT)*(TIA_table$bwgt_diff))
ATT
```

```{r}

model3 <- lm_robust(birthwgt ~ tobacco + 
                      mage + meduc + mblack + alcohol + 
                      mage:meduc + mage:mblack + mage:alcohol +
                      meduc:mblack + meduc:alcohol + mblack:alcohol +
                      mage:meduc:mblack + mage:meduc:alcohol + meduc:mblack:alcohol +
                      mage:meduc:mblack:alcohol, data = data_matching)

summary(model3)

```

(d) Estimate the propensity score for maternal smoking using a logit estimator and based on the following specification: mother’s age, mother’s age squared, mother’s education, and indicators for mother’s race, and alcohol consumption. 


```{r}
# this is the model of the propensity score 
ps_model <- glm(tobacco ~ mage + mage_sq + meduc + mblack + alcohol, family = binomial(), data = data_matching)
summary(ps_model)

EPS <- predict(ps_model, type = "response")
PS_WGT <- (data_matching$tobacco/EPS) + ((1-data_matching$tobacco)/(1-EPS))

```

(e) Use the propensity score weighted regression (WLS) to estimate the effect of maternal smoking on birth weight (Lecture 7, slide 12).

```{r}

# WLS USING EPS WEIGHTS
wls1 <- lm(birthwgt ~ tobacco, data_matching, weights=PS_WGT)
summary(wls1)
se_model = starprep(wls1, stat = c("std.error"), se_type = "HC2", alpha = 0.05)
stargazer(wls1, se = se_model, type="text")
```

